---
title: "Working with ellmer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with ellmer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

`send_prompt()` accepts an ellmer Chat directly or an `llm_provider_ellmer()`.
Pin the model when creating the chat if reproducibility matters; provider
defaults can change. The adapter preserves native turns, content, usage, cost
and citations in the full result.

## Callbacks and chat ownership

Each `send_prompt()` evaluation deep-clones its native chat. Callback registries,
tools and conversation state on that working chat are independent of the original.
A callback's R closure still refers to the variables it originally captured.
Register callbacks that modify history on the working chat, inside a prompt
wrap's `parameter_fn`, which runs after cloning:

```{r}
library(tidyprompt)
chat <- ellmer::chat_openai(model = "gpt-4.1-mini")
prompt <- prompt_wrap("Summarize the discussion", parameter_fn = function(llm_provider) {
  working <- llm_provider$get_chat()
  working$on_request_start(function(turns) {
    # Inspect `turns` and, if needed, call working$set_turns(...).
    message("Starting a model request")
  })
  list()
})
result <- send_prompt(prompt, chat, return_mode = "full")
result$ellmer_chat$get_turns()
```

Callbacks require the corresponding methods in your ellmer version. The
adapter honors history replacement performed by request callbacks, including
compaction, when preparing later feedback turns.
