---
title: "Working with ellmer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with ellmer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

`send_prompt()` accepts an ellmer Chat directly or an `llm_provider_ellmer()`.
Pin the model when creating the chat if reproducibility matters; provider
defaults can change. The adapter preserves native turns, content, usage, cost
and citations in the full result.

| Capability | Support |
|---|---|
| Direct Chat input, native tools and structured output | ellmer >= 0.3.0 |
| Built-in tools and ignored tool arguments | ellmer >= 0.4.0 |
| Native history, thinking, usage and content preservation | Preserved when the installed ellmer exposes them |
| Generic attachments | `add_content()`; content classes and provider support vary |
| Structured streaming | ellmer 0.5.0 with native provider support; blocking fallback otherwise |
| Rich stream events and cancellation | ellmer 0.5.0 |
| Request hooks and `limit_requests()` with an 'ellmer' provider | 'ellmer' 0.5.0 |
| Context-aware tools | Native ellmer execution only |
| Native async, parallel and batch evaluation | Not exposed by `send_prompt()` |

## Callbacks and chat ownership

Each `send_prompt()` evaluation deep-clones its native chat. Callback registries,
tools and conversation state on that working chat are independent of the original.
A callback's R closure still refers to the variables it originally captured.
Register callbacks that modify history on the working chat, inside a prompt
wrap's `parameter_fn`, which runs after cloning:

```{r}
library(tidyprompt)
chat <- ellmer::chat_openai(model = "gpt-4.1-mini")
prompt <- prompt_wrap("Summarize the discussion", parameter_fn = function(llm_provider) {
  working <- llm_provider$get_chat()
  working$on_request_start(function(turns) {
    # Inspect `turns` and, if needed, call working$set_turns(...).
    message("Starting a model request")
  })
  list()
})
result <- send_prompt(prompt, chat, return_mode = "full")
result$ellmer_chat$get_turns()
```

Callbacks require the corresponding methods in your ellmer version. The
adapter honors history replacement performed by request callbacks, including
compaction, when preparing later feedback turns.

## Schemas and tool results

Closed objects use ellmer's native types. Open objects and constraints such as
numeric bounds or schema composition use `ellmer::type_from_schema()` to preserve
the schema. Provider/model support for these schemas still varies. Raw schemas
return JSON-shaped R values rather than ellmer's typed factors and data frames.
Install `jsonvalidate` to validate constraints locally; tidyprompt requires it
when a schema has constraints beyond the basic native types.

Tidyprompt-created tools serialize lists as JSON objects/arrays and data frames
as arrays of row objects. Return an explicit JSON string to control serialization.
Native ellmer tools retain their own result contract. Rich content results and
tools using `ellmer::tool_context()` need native ellmer execution. Cross-provider
tool conversion preserves nested argument schemas and the native `convert` policy.

## Documents and uploaded files

Use `add_content()` for any native ellmer Content object, including documents,
PDFs and uploaded-file references. Ellmer remains responsible for file uploads,
expiry and provider restrictions. Attachments work with structured output and
streaming, and stay in native history during feedback without being reattached.

```{r}
document <- ellmer::content_document_file("report.txt")
result <- add_content("Summarize this document", document) |>
  send_prompt(chat, return_mode = "full")
```

## Structured streams

With ellmer 0.5.0, `answer_as_json()` and `stream = TRUE` stream raw JSON chunks
through the usual `stream_callback(chunk, meta)`. Final extraction keeps ellmer's
R coercion, including data frames and factors. Providers/models that use
tool-based structured output, older ellmer versions, and custom adapters without
the required capabilities use blocking `chat_structured()` instead.

## Stream events and recovery

Set provider parameter `stream_content = TRUE` to receive native events in
`meta$content`. The first callback argument remains a string: non-text events
such as citations, thinking and tool results receive `""` and do not enter the
accumulated answer. Inspect `meta$content_type` to render these separately.

```{r}
controller <- ellmer::stream_controller()
provider <- llm_provider_ellmer(chat, parameters = list(
  stream = TRUE, stream_content = TRUE, stream_controller = controller
))
provider$stream_callback <- function(chunk, meta) {
  cat(chunk)
  # Call controller$cancel() to stop after the next chunk boundary.
}
result <- tryCatch(
  send_prompt("Explain the report", provider),
  tidyprompt_stream_error = function(e) {
    # Also catches tidyprompt_stream_cancelled, a subclass.
    list(partial = e$partial_response, chat = e$ellmer_chat, turn = e$partial_turn)
  }
)
```

Cancellation and iteration failures stop validation and feedback requests.
The condition exposes the working chat and ellmer's partial turn; partial text
is never treated as a successful answer. Iteration failures do not trigger a
second model request. Controllers require a streaming-capable path; the
blocking structured fallback cannot be cancelled with a stream controller.

## Request limits and observability

`max_interactions` limits tidyprompt's outer evaluation loop. Use
`limit_requests(prompt, max_requests = 5)` to bound individual model requests
across tool calls and feedback for regular tidyprompt providers and 'ellmer'.
With an 'ellmer' provider, this requires 'ellmer' 0.5.0 and counts the blocking
structured path explicitly, because that method bypasses its request hooks.
The limit stops before the next model request and raises
`tidyprompt_request_limit` with the working provider (and native chat when
applicable) attached. It does not prevent
execution of tools requested by an already completed model request, or bound
transport-level retries made within one model request.

```{r}
prompt <- limit_requests("Research this question", 5)
prompt <- prompt_wrap(prompt, parameter_fn = function(llm_provider) {
  working <- llm_provider$get_chat()
  working$conversation_id <- "research-session-123"
  working$on_request_end(function(turn) {
    print(turn@tokens)
    print(turn@cost)
  })
  list()
})
result <- send_prompt(prompt, chat, return_mode = "full")
result$ellmer_chat$get_tokens()
```

Use native turn metadata and ellmer tracing for usage and latency; the adapter's
HTTP fields do not represent native requests. `on_request_end` itself does not
fire on the blocking structured path in ellmer 0.5.0; read the returned native
turns there. Token estimates and recorded cost cannot guarantee a final spending
cap, because a request may exceed the remaining estimate before it completes.

## Tools followed by structured extraction

Ellmer suppresses tools during native structured extraction. Use two evaluations
when a task needs both: first gather information using tools, then extract a
schema-constrained answer from that conversation.

```{r}
research <- answer_using_tools("Find the relevant facts", tools = my_tools,
  type = "ellmer") |>
  send_prompt(chat, return_mode = "full")

answer <- add_msg_to_chat_history(research$chat_history, "Extract the final result") |>
  answer_as_json(schema = ellmer::type_object(summary = ellmer::type_string()),
    type = "ellmer") |>
  send_prompt(llm_provider_ellmer(research$ellmer_chat))
```

## Async, parallel and batch boundaries

`send_prompt()` is synchronous, including its extraction, validation and feedback
loop. Calling it from an ellmer async or batch helper does not make that loop
asynchronous. For Shiny, the separate-process approach in
`vignette("streaming_shiny_ipc")` remains available. Construct each chat in its
worker instead of sharing a mutable Chat across concurrent evaluations.

Use ellmer's native async/parallel/batch APIs directly for independent native
requests that do not need tidyprompt's evaluation loop. A future integrated path
must define asynchronous validation and tools, cancellation, and scheduling of
data-dependent retries; those capabilities are not implied by this adapter.
